Novel lattice Boltzmann method for simulation of strongly shear thinning viscoelastic fluids

### 2024-02-13 NovelLBM.tex
First re-read

The simulation of viscoelastic liquids using the Lattice-Boltzmann method in full three dimensions remains a formidable numerical challenge.
In particular the simulation of strongly shear-thinnig fluids, where the ratio between the high-shear and low-shear viscosities is large, is often prevented by stability problems.
Here we present a novel approach to overcome these issues.
The central idea is to artificially increase the solvent viscosity which allows the method to benefit from the very good stability properties of the LBM.
To compensate for this additional viscous stress, we reduce the polymer stress by the same amount.
We apply this novel method to simulate two realistic cell carrier liquids, methyl cellulose and alginate solutions, of which the latter exhibits a viscosity ratio of more than 10,000.

@@

Microfluidic technoglogies involving living cells are becoming increasingly popular. 
Examples include medical diagnostics, where the pathological alteration of cell properties can serve as a marker for disease detection \cite{Toepfner.2018} or biofabrication where living cells are printed to create artificial tissues \cite{possl2021, Poologasundarampillai.2021, oconnell2022}. 
In all these applications cells are suspended in a cell carrier fluid which typically contains dissolved polymers making the fluid highly viscoelastic. 
The motivation behind using viscoelastic fluids differs between applications. 
In medical diagnostics, the viscous stress must be large enough to achieve detectable cell deformation, while in biofabrication the printed construct must be stable enough to retain its shape after the printing process. 
To satisfy the latter requirement, cell carriers used in biofabrication („bioinks“) typically exhibit a very high viscosity ratio with the zero-shear viscosity being many orders of magnitude larger than the pure solvent viscosity \cite{paxton2017}.

To understand the rheology of such complex liquids in non-trivial geometries, computer simulations are an invaluable tool. 
While a number of numerical techniques have been developed, the simulation of viscoelastic liquids remain an involved subject \cite{Alves.2020}, in particular for high viscosity ratios. 
A very popular simulation technique for Newtonian fluids is the Lattice Boltzmann method \cite{Kruger_book} due to its simplicity and high parallelizability. 
LBM has also been extended to viscoelastic situations, mainly in two \cite{ispolatov2002,wang2019,su2013,giraud1998,xie2018,dzanic2022,frank2006,zou2014,lee2017,dzanic2022_2,sedaghat2021,frantziskonis2011,qin2023,yoshino2008,papenkort2015,phillips2011,onishi2005} but occassionally also in three \cite{malaspinas2010,gupta2015,lallemand2003,ma2020,onishi2006,wang2020,dellar2014,su2013_2,wang2021,kuron2021,gupta2016} dimensions.
To alleviate stability problems, some of these works include an artificial diffusivity of the polymer stress \cite{malaspinas2010,ma2020,wang2020,wang2021}. 
Other remedies include special decompositions of the stress tensor\cite{dzanic2022} or optimization of the solvent contribution\cite{gupta2015}. 
So far, however, none of these methods has been applied to very high viscosity ratio fluids which are often used as cell carrier fluids in lab experiments or technological applications.

@@

%	For a solution to be well suited for usage as a bioink during bioprinting its rheology has to fulfill certain requirements. During the printing process it should flow easily through the nozzle, meaning the viscosity should be low. This is not only beneficial for ease of use, but is also crucial for survivability of the cells, as the stress they experience is dependent on the viscosity. However, after the printing process, it is desired, that the printed construct is stable. This can be temporarily archived by very high viscosities. Therefore, one approach to solve this is to use shear thinning fluids. Shear thinning fluids can easily be produced by dissolving polymers in water. This however, has the side effect of viscoelasticity. Due to the polymers the solution also exhibits elastic behavior. The influence of which on the printing process is not well understood yet. For realistic bioinks, the ratio of their zero shear viscosity and their infinite shear viscosity is large as is expected based on the desired attributes mentioned above. For such large viscosity ratios, simulating viscoelastic materials with Lattice Boltzmann methods was to our knowledge not yet possible.

%	Such simulations would however be very beneficial. They allow the \glqq measurement\grqq{ }of parameters that are otherwise hard or even impossible to access. They allow the prediction of the outcome of experiments without performing them, which is especially helpful if the experiment cannot be easily performed. An example would be the design of new and improved printing nozzles. With simulations, the geometry can be optimized to a degree, that would be near impossible using iterative experiments. The large amounts of data simulations can provide allow better understanding of the fundamental processes, that cause observed behavior. \todo{"List some simulations for bionprinting, bioinks that have been done (restrict to only viscoelastic sims)"}

%	Simulations using Lattice Boltzmann methods (LBM) are popular, as it has many advantages. It is very extensible and thus can deal with complex geometries, particles (e.g. cells) suspended in the fluid and many different interactions within and between those. Most importantly it is and easily parallelizable algorithm and thus can be run on GPUs or similar accelerators. This allows even large domains to be simulated in short amount of time with commodity GPUs and no need for any supercomputer. It is also easily possible to extend LBM to deal with viscoelastic fluids. This has been done before by multiple groups. There are many publications discussing such approaches in two dimensions\cite{ispolatov2002,wang2019,su2013,giraud1998,xie2018,dzanic2022,frank2006,zou2014,lee2017,dzanic2022_2,sedaghat2021,frantziskonis2011,qin2023,yoshino2008,papenkort2015,phillips2011,onishi2005}. 		Simulations in three dimensions have a higher computational cost and are not as abundant, but have also been performed by many\cite{malaspinas2010,gupta2015,lallemand2003,ma2020,onishi2006,wang2020,dellar2014,su2013_2,wang2021,kuron2021,gupta2016}. Among these many are concerned about stability. Some include diffusivity of the polymer stress in their model to increase stability\cite{malaspinas2010,ma2020,wang2020,wang2021}. Other remedies include special decompositions of the stress tensor\cite{dzanic2022} or optimization of the solvent contribution\cite{gupta2015}. However, none of these reach the relevant viscosity ratios for bioinks.

%	Here we show our novel simulation approach for bioinks. It is specifically tailored to support arbitrary viscosity ratios and thus is able to simulate common bioinks or any other viscoelastic material with similar properties while providing high accuracy and stability. This is archived by subtracting a viscous stress from the polymer stress calculated from the PTT model. This offset is corrected by increasing the viscosity of the base LBM algorithm accordingly.

@@

To illustrate our novel LBM algorithm, we consider two prototypical cell carrier fluids.
The first is an alginate solution which is often used in biofabrication, i.e., the 3D printing of living cells, but can also serve as a common food additive.
Our solution contains 4\% alginate solvated in PBS as was measured in a commerical plate-plate rheometer (see appendix  XX for experimental details).
The shear-dependent viscosity $ \eta$ and the first normal stress difference $N_1$ are shown as function of shear rate in a frequency-sweep \SG{ ? } measurement in figure~\ref{fig:fit_alginate}.
We note the very high zero-shear viscosity of more than $10000 \si{ mPa \cdot s}$.
Carrier fluids in bioprinting, termed bioinks, are characterized in general by such high viscosities to guarantee stability of the printed construct.
On the other hand, during the actual printing process when the material is flowing, the viscosity should be as low as possible to avoid excessive mechanical stress on the embedded cells.
	\begin{figure}[H]
		\centering
		\includegraphics[width=7.8cm]{assets/plots/fit_alginate_eta.eps}
		\includegraphics[width=7.8cm]{assets/plots/fit_alginate_N.eps}
%		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/fit_alginate_eta.eps}
%		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/fit_alginate_N.eps}
		\caption{Rheology data for alginate solution from 21 measurements of the viscosity (a) and normal stress (b).
				Orange lines show a fit of the PTT rheological model. \SG{ larger font in figures }}
		\label{fig:fit_alginate}
	\end{figure}

@@

Due to the complexity of viscoelastic fluids, there is an abundance of (semi-)empirical models to describe rheological data.
Common options include the Oldroyd-B \cite{oldroyd1958} and the FENE-P model \cite{bird1980,bird1987}.
The Oldroyd-B model is not appropriate for most real complex fluids as it features a shear-independent viscosity in clear contradiction to the experimental data presented in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc}.
Although the FENE-P does reproduce shear-thinning, we were not able to obtain a satisfactory fit to our data for alginate and methyl cellulose with physically reasonable values of the model parameters.
The best fit was obtained with the so-called PTT model due to Phan-Thien and Tanner \cite{phan-thien1977,phan-thien1978} which was for explicitly derived for polymer networks in solution and thus appears to capture well the essential physics behind the alginate solution. \SG{ check that these papers have the exponential version }.

The PTT model is specified by a constitutive equation for the polymer stress $ \ma{\tau}$:
	\begin{equation}
		\overset{\nabla}{\ma{\tau}} = -\xi\left(\ma{\tau}\cdot\ma{D} + \ma{D}\cdot\ma{\tau}\right) - e^{\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}}\frac{\ma{\tau}}{\lambda} + 2\frac{\eta_\text{p}}{\lambda}\ma{D}
		\label{eq:polymerStress}
	\end{equation}
where $\ma{D}$ is the strain rate tensor. 
The polymer viscosity is denoted $\eta_\text{p}$ and $\lambda$ is the corresponding relaxation time.  \SG{ are they actually independent? }
$\epsilon$ is the so-called extensibility parameter which controls at which shear rates thinning takes place and $\xi$ describes the slip in the polymer network which we will set to zero in the following.
The upper convected derivative is defined as follows:
	\begin{equation}
		\overset{\nabla}{\ma{A}} = \frac{\Dif\ma{A}}{\Dif t} - \left(\left(\nabla\vec{u}\right)^\text{T}\cdot\ma{A} + \ma{A}\cdot\left(\nabla\vec{u}\right)\right)
	\end{equation}
where $\vec{u}$ is the fluid velocity and $\left(\nabla\vec{u}\right)_{ij} = \nabla_i\vec{u}_j$. 
The material derivative is
	\begin{equation}
		\frac{\Dif\ma{A}}{\Dif t} = \frac{\partial\ma{A}}{\partial t} + \vec{u}\cdot\nabla\ma{A}.
	\end{equation}	
Finally, the total fluid stress is the sum of the polymer stress $ \tau $ and a solvent stress $ \tau _ \mathrm{s}$ due to the Newtonian solvent with viscosity $ \eta _ \mathrm{s}$.

For a simple shear flow in the $xy$-plane, an analytical expression for $ \eta $ and $N_1$ is given by \cite{ferras2019} which (with $\xi=0$) reads
	\begin{equation}
		\eta\left(\dot{\gamma}\right) = \frac{\eta_\text{p}}{\exp\left[0.5W_0\left(4\epsilon \mathrm{Wi}^2\right)\right]} + \eta _\mathrm{s} \ma{D} \label{eq:shearEta}
	\end{equation}
	\begin{equation}
		N_1 = \frac{\eta_\text{p}}{2\epsilon\lambda}W_0\left(4\epsilon \mathrm{Wi}^2\right) \label{eq:shearN}
	\end{equation}
where $ \mathrm{Wi} = \lambda \dot{ \gamma  }$ is the Weissenberg number and $W_0$ is the Lambert $W$-function. \SG{ check that $ \eta _ \mathrm{s}$ is correct }
\SG{ I did not find these equations in \cite{ferras2019}? }
Fitting these equations to the rheological data for alginate and methy cellulose leads to satisfactory agreement as shown in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc}.
In the fit, the solvent viscosity was fixed at $ \eta _ \mathrm{s} = 1 \si{ mPa } \cdot \si{ s }$ as the viscosity of the PBS buffer. \SG{ give ref }
At very high shear rates, the contained polymers are completely stretched and the viscosity asymptotically approaches the solvent values $ \eta _ \mathrm{s}$.
The obtained parameters for alginate are
	\begin{align}
		\eta_\text{p} &= \SI{48.2\pm0.4}{\pascal\second}\\
		\lambda &= \SI{0.343 \pm 0.004}{\second}\\
		\epsilon &= \num{0.545 \pm 0.010}.
	\end{align}
while the parameters for the three concentrations of methyl cellulose are given in table~\ref{tab:MC_parameters}.
	\begin{table}[H]
		\centering
		\begin{tabular}{c|c|c|c}
%			Concentration/\unit{\percent} & $\eta_\text{p}/\unit{\milli\pascal\second}$ & $\lambda/\unit{\milli\second}$ & $\epsilon$ \\
			Concentration/ percent & $\eta_\text{p}/mPas$ & $\lambda/ms$ & $\epsilon$ \\
			\hline
			\num{0.49} & \num{18.7\pm0.4} & \num{0.344\pm0.003} & \num{0.270\pm0.003} \\
			\num{0.59} & \num{32.5\pm0.7} & \num{0.433\pm0.004} & \num{0.365\pm0.004} \\
			\num{0.83} & \num{81\pm2} & \num{0.714\pm0.009} & \num{0.496\pm0.006}
		\end{tabular}
		\caption{Parameters of the PTT modelobtained when fitting the methyl cellulose solutions.}
		\label{tab:MC_parameters}
	\end{table}

We introduce the ratio between the zero-shear and the infinite-shear (i.e.\ solvent) viscosity as 
\begin{align}
	R_ \eta  = \frac{ \eta _ \mathrm{p} } { \eta _ \mathrm{s} }		\label{eq:R_eta}
\end{align}
which will turn out to be an important parameter to determine numerical stability of LB simulations. \SG{ do we actually need a ratio? Since $ \eta _\mathrm{s}$ is probably always water-like, would it not be enough to state whether $ \eta _ \mathrm{p}$ is high or low } \SG{ do we actually show that $R_ \eta  $ determines stability? }
This ratio can in a similar fashion be defined for other models such as FENE-P or the inelastic, but shear-thinning Carreau-Yasuda model \cite{carreau1972, yasuda1979}.
For our alginate solution we obtain $R _ \eta = \num{48.2\pm0.4e3}$, indicating that the solution viscosity at very low shear rates is more than four orders of magnitude larger than the solvent viscosity which represents a severe challenge to standard numerical algorithms.
That this characteristic is not unique to alginate, but is in fact a generic feature of cell carrier fluids used in bioprinting is shown in table~\ref{tab:bioinks} where rheological from various literature sources have been compiled into a single format to determine $ R_ \eta $.
The 0.49 \% methyl cellulose solution used in RT-DC measurements, in contrast, features a much lower value of $R_\eta = \num{18.7\pm0.4}$.
\begin{table}
	\centering
		\begin{tabular}{l|p{3cm}|p{1.5cm}|p{1.5cm}|p{1.5cm}|p{2cm}}
%			Author & Material & $\eta_0/\unit{\pascal\second}$ & $\eta_\infty/\unit{\pascal\second}$ & $R_\eta$ & extracted from \\
			Author & Material & $\eta_0/Pas$ & $\eta_\infty/Pas$ & $R_\eta$ & extracted from \\
			\hline
			Amorim\cite{amorim2021} & pre-crosslinked alginate &  >1e4 & <3e-1 & >3e4 & Fig. 3b \\
			\rowcolor{secondaryBackground}
			Pössl\cite{possl2021} & blend & 3e2 & 1e-4 & 3e6 & Table 4 \\
%			Paxton\cite{paxton2017} &Nivea Creme\n Alginate& 1e5\n5e4 & <1\n<5e1 & >1e5\n>1e3 & Fig. 4b \\
			Paxton\cite{paxton2017} & Alginate& 5e4 & <5e1 & >1e3 & Fig. 4b \\
			\rowcolor{secondaryBackground}
			O'Connell\cite{oconnell2020} & pluronic F-127 & >1e6 & <8 & > 2e5 & Chapter 7 Fig. 13 \\
		\end{tabular}
	\caption{Compilation of literature data on typical bioinks showing that a very viscosity ratio $ R_ \eta $ is a generic feature of these liquids. }
	\label{tab:bioinks}
\end{table}

@@

\SG{ make sure that all papers cited in the introduction as ''3D'' are discussed here as well }
\SG{ and vice versa: make sure that all papers discussed here are also cited in the introduction }
Since the early days of LBM methods for viscoelastic fluids have been developed. 
Keeping in mind our target geometries, we focus here on three dimensional formulations.
The first approaches avoided an explicit modeling of a viscoelastic constitutive equation and simply modified the LB collision operator to obtain a fluid obeying Jeffery's viscoelastic model \cite{lallemand2003}. \SG{ can we say more about Jeffery? Why is it bad? Is it shear-thinning? }
More advcanced approaches include the work of Onishi \textit{et al.} \cite{onishi2005, onishi2006} who considered a non-shear-thinning Oldroyd-B model using a microscopic Fokker-Planck equation for the polymer dumbbells. 
The work of \cite{malaspinas2010} and \cite{ma2020} considered in addition the FENE-P model solving the advection part of the model using a second LBM running in parallel to the Navier-Stokes LBM while \cite{wang2020} used this approach for the Oldroyd-B model.
Later, hybrid schemes coupling an LB solver for the solvent to a finite-difference \cite{gupta2015_2, gupta2015, gupta2016} or finite-volume \cite{kuron2021} scheme for the polymer stress (or, alternatively, the polymer conformation) tensor were introduced.
To improve numerical stability, some authors propose the inclusion of an artificial diffusivity \SG{ cite }, which introduces a small error compared to the analytical solution of the constitutive equation. 
Coupling of the resulting polymer stress to the Navier-Stokes LBM can be achieved via the inclusion of point forces or via a modification of the equilibrium populations.
A recent comparison in 2D found that the former in general leads to higher stability \cite{dzanic2022}.

The stable range of Weissenberg numbers covered by existing methods is up to $ \mathrm{Wi}  \approx  10$ for the method of \cite{malaspinas2010} and up to $ \mathrm{Wi}  \approx 100$ in \cite{gupta2015,ma2020,gupta2016}.
However, another stability criterion, which is often overlooked is the viscosity ratio $ R_ \eta $ as defined in XX. 
Existing methods often stay at or below $R_\eta = 1$\cite{gupta2015,onishi2006,wang2020,su2013_2,gupta2015_2}, while the ones trying higher ratios report being limited below $R_\eta = 10$\cite{malaspinas2010,ma2020,kuron2021} due to stability issues. 
The details are listed in table~\ref{tab:viscRatio}.
Comparing to the rheological data presented in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc}, we conclude that existing LB methods appear to be appropriate only for relatively dilute solutions and in particular do not cover the viscosity ratio required by these technologically important cell carrier fluids.
We therefore set out to develop a novel viscoelastic LB scheme which we will present in the next section.

@@

The Lattice-Boltzmann method (LBM) has become a standard tool in numerical fluid mechanics \cite{Kruger_book} and we therefore describe it here only very briefly.
We use a D3Q19 lattice in full three dimensions together with a single relaxation time collission operator.
Our implementation is based on FluidX3D which has been extensively used and validated in earlier publications \cite{Lehmann.20229da}.
The core of the algorithm is implemented in OpenCL and thus capable of running on GPUs from different vendors.
Boundaries are modeled using the half-way bounce back algorithm. 
\SG{ and inflow/outflow: give details }

Physical units are mapped to lattice units to obtain a dimensionless relaxation time of 1.
The employed lattice dimensions vary for the different applications and are specified below. \SG{ to do }
To drive the flow, we use either a constant body force (modeling a physical pressure gradient) or inflow-outflow boundaries as specified in the different application scenarios below. \SG{ check, specify }

@@

\section{Rheology measurements of alginate solutions}

>>>>

The simulation of viscoelastic liquids using the Lattice-Boltzmann method in full three dimensions remains a formidable numerical challenge.
In particular the simulation of strongly shear-thinnig fluids, where the ratio between the high-shear and low-shear viscosities is large, is often prevented by stability problems.
Here we present a novel approach to overcome these issues.
The central idea is to artificially increase the solvent viscosity which allows the method to benefit from the very good stability properties of the LBM.
To compensate for this additional viscous stress, we reduce the polymer stress by the same amount.
We apply this novel method to simulate two realistic cell carrier fluids, methyl cellulose and alginate solutions, of which the latter exhibits a viscosity ratio exceeding 10,000.

@@

Microfluidic technoglogies involving living cells are becoming increasingly popular. 
Examples include medical diagnostics, where the pathological alteration of cell properties can serve as a marker for disease detection \cite{Toepfner.2018} or biofabrication where living cells are printed to create artificial tissues \cite{possl2021, Poologasundarampillai.2021, oconnell2022}. 
In all these applications cells are suspended in a cell carrier fluid which typically contains dissolved polymers making the fluid highly viscoelastic and shear-thinning. 
The motivation behind using viscoelastic fluids differs between applications. 
In medical diagnostics, the viscous stress must be large enough to achieve detectable cell deformation, while in biofabrication the printed construct must be stable enough to retain its shape after the printing process. 
To satisfy the latter requirement, cell carriers used in biofabrication („bioinks“) typically exhibit a very high viscosity ratio with the zero-shear viscosity being many orders of magnitude larger than the pure solvent viscosity \cite{paxton2017}.

To understand the rheology of such complex liquids in non-trivial geometries, computer simulations are an invaluable tool. 
While a number of numerical techniques have been developed, the simulation of viscoelastic liquids remain an involved subject \cite{Alves.2020}, in particular for high viscosity ratios. 
A very popular simulation technique for Newtonian fluids is the Lattice Boltzmann method \cite{Kruger_book} due to its simplicity and high parallelizability. 
LBM has also been extended to viscoelastic situations, mainly in two \cite{ispolatov2002,wang2019,su2013,giraud1998,xie2018,dzanic2022,frank2006,zou2014,lee2017,dzanic2022_2,sedaghat2021,frantziskonis2011,qin2023,yoshino2008,papenkort2015,phillips2011,onishi2005} but occassionally also in three \cite{malaspinas2010,gupta2015,lallemand2003,ma2020,onishi2006,wang2020,dellar2014,su2013_2,wang2021,kuron2021,gupta2016} dimensions.
To alleviate stability problems, some of these works include an artificial diffusivity of the polymer stress \cite{malaspinas2010,ma2020,wang2020,wang2021}. 
Other remedies include special decompositions of the stress tensor\cite{dzanic2022} or optimization of the solvent contribution\cite{gupta2015} \SG{ what is this? }.  
Despite these improvements, however, none of these methods has been applied to very high viscosity ratio fluids which are often used as cell carrier fluids in lab experiments or technological applications.

@@

@@

To illustrate our novel LBM algorithm, we consider two prototypical cell carrier fluids.
The first is an alginate solution which is often used in biofabrication but can also serve as a common food additive.
Our solution contains 4\% alginate solvated in PBS and was measured in a commerical plate-plate rheometer (see appendix  \ref{app:experiment} for experimental details).
The shear-dependent viscosity $ \eta$ and the first normal stress difference $N_1$ are shown as function of shear rate in figure~\ref{fig:fit_alginate}.
We note the very high zero-shear viscosity of more than $10000 \si{ mPa \cdot s}$.
Bioinks in biofabrication are characterized in general by such high viscosities to guarantee stability of the printed construct.
At the same time, during the actual printing process when the material is flowing, the viscosity should be as low as possible to avoid excessive mechanical stress on the embedded cells.
This motivates the use of highly shear-thinning viscoelastic fluids in biofabrication.
	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=0.5]{assets/plots/fit_alginate_eta.eps}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=0.5]{assets/plots/fit_alginate_N.eps}
		\caption{Rheology data for alginate solution from 21 measurements of the viscosity (a) and normal stress (b).
				Orange lines show a fit of the PTT rheological model. \SG{ larger font in figures }}
		\label{fig:fit_alginate}
	\end{figure}

@@

Due to the complexity of viscoelastic fluids, there is an abundance of (semi-)empirical models to describe rheological data.
Common options include the Oldroyd-B \cite{oldroyd1958} and the FENE-P model \cite{bird1980,bird1987}.
The Oldroyd-B model is not appropriate for most real complex fluids as it features a shear-independent viscosity in clear contradiction to the experimental data presented in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc}.
%Although the FENE-P does reproduce shear-thinning, we were not able to obtain a satisfactory fit to our data for alginate and methyl cellulose with physically reasonable values of the model parameters.
A good fit to our data was obtained with the so-called PTT model due to Phan-Thien and Tanner \cite{phan-thien1977,phan-thien1978} which has been developed explicitly for polymer networks in solution and thus appears to capture well the essential physics behind the alginate and methyl cellulose solutions.

The PTT model is specified by a differential equation for the polymer stress $ \ma{\tau}$:
	\begin{equation}
		\overset{\nabla}{\ma{\tau}} = -\xi\left(\ma{\tau}\cdot\ma{D} + \ma{D}\cdot\ma{\tau}\right) - e^{\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}}\frac{\ma{\tau}}{\lambda} + 2\frac{\eta_\text{p}}{\lambda}\ma{D}
		\label{eq:polymerStress}
	\end{equation}
where $\ma{D}$ is the strain rate tensor. 
\SG{ check that PTT is given in this form in \cite{phan-thien1977,phan-thien1978} or cite other source }
The polymer viscosity is denoted $\eta_\text{p}$ and $\lambda$ is the corresponding relaxation time.  
$\epsilon$ is the so-called extensibility parameter which controls at which shear rates thinning takes place and $\xi$ describes the slip in the polymer network which we will set to zero in the following.
The upper convected derivative of an arbitrary tensor $ \ma{ A }$ is defined as follows:
	\begin{equation}
		\overset{\nabla}{\ma{A}} = \frac{\Dif\ma{A}}{\Dif t} - \left(\left(\nabla\vec{u}\right)^\text{T}\cdot\ma{A} + \ma{A}\cdot\left(\nabla\vec{u}\right)\right)
	\end{equation}
where $\vec{u}$ is the fluid velocity and $\left(\nabla\vec{u}\right)_{ij} = \nabla_i\vec{u}_j$. 
The material derivative is
	\begin{equation}
		\frac{\Dif\ma{A}}{\Dif t} = \frac{\partial\ma{A}}{\partial t} + \vec{u}\cdot\nabla\ma{A}.
	\end{equation}	
Finally, the total fluid stress $\ma{\sigma}$ is the sum of the polymer stress $ \tau $ and a solvent stress due to the Newtonian solvent with viscosity $ \eta _ \mathrm{s}$:
	\begin{align}
		\ma{\sigma} &=  \ma{\tau} + 2\eta_\text{s}\ma{D} 		\label{eq:totalStress}
	\end{align}

For a simple shear flow in the $xy$-plane of a PTT fluid with $ \xi =0$, an analytical expression for $ \eta $ and $N_1$ can be derived as detailed in appendix \ref{app:semianalyticalPoiseuille}:
	\begin{equation}
		\eta\left(\dot{\gamma}\right) = \frac{\eta_\text{p}}{\exp\left[0.5W_0\left(4\epsilon \mathrm{Wi}^2\right)\right]} + \eta _\mathrm{s} \label{eq:shearEta}
	\end{equation}
	\begin{equation}
		N_1 = \frac{\eta_\text{p}}{2\epsilon\lambda}W_0\left(4\epsilon \mathrm{Wi}^2\right) \label{eq:shearN}
	\end{equation}
where $ \mathrm{Wi} = \lambda \dot{ \gamma  }$ is the Weissenberg number and $W_0$ is the Lambert $W$-function. \SG{ check that $ \eta _ \mathrm{s}$ is correct }
Fitting these equations to the rheological data for alginate and methy cellulose leads to satisfactory agreement as shown in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc}.
In the fit, the solvent viscosity was fixed at $ \eta _ \mathrm{s} = 1 \si{ mPa } \cdot \si{ s }$ as the viscosity of the PBS buffer. \SG{ give ref }
%At very high shear rates, the contained polymers are completely stretched and the viscosity asymptotically approaches the solvent value $ \eta _ \mathrm{s}$.
The obtained parameters are given in table~\ref{tab:PTTparameters}.
	\begin{table}[H]
		\centering
		\begin{tabular}{c|c|c|c}
%			Concentration/\unit{\percent} & $\eta_\text{p}/\unit{\milli\pascal\second}$ & $\lambda/\unit{\milli\second}$ & $\epsilon$ \\
			Fluid & $\eta_\text{p}/mPas$ & $\lambda/ms$ & $\epsilon$ \\
			\hline
			alginate \num{4} & \num{48,200\pm0.4} & \num{343 \pm 0.004}. & \num{0.545 \pm 0.010}\\
			methyl c. \num{0.49} & \num{18.7\pm0.4} & \num{0.344\pm0.003} & \num{0.270\pm0.003} \\
			methyl c. \num{0.59} & \num{32.5\pm0.7} & \num{0.433\pm0.004} & \num{0.365\pm0.004} \\
			methyl c. \num{0.83} & \num{81\pm2} & \num{0.714\pm0.009} & \num{0.496\pm0.006}
		\end{tabular}
		\caption{Parameters of the PTT model obtained by fitting to the rheological measurements in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc}.
		\SG{ add a column with $R_ \eta $ }}
		\label{tab:PTTparameters}
	\end{table}

We introduce the ratio between the zero-shear and the infinite-shear (i.e.\ solvent) viscosity as 
\begin{align}
	R_ \eta  = \frac{ \eta _ \mathrm{p} } { \eta _ \mathrm{s} }		\label{eq:R_eta}
\end{align}
which will turn out to be an important parameter to determine numerical stability of LB simulations. 
\SG{ wo don't actually show that $R_ \eta  $ determines stability? }
This ratio can in a similar fashion be defined for other models such as FENE-P or the inelastic, but shear-thinning Carreau-Yasuda model \cite{carreau1972, yasuda1979}.
For our alginate solution we obtain $R _ \eta = \num{48.2\pm0.4e3}$, indicating that the solution viscosity at very low shear rates is more than four orders of magnitude larger than the solvent viscosity which represents a severe challenge to standard numerical algorithms.
That this characteristic is not unique to alginate, but is in fact a generic feature of cell carrier fluids used in bioprinting is shown in table~\ref{tab:bioinks} where rheological data from various literature sources have been compiled into a single format to determine $ R_ \eta $.
The 0.49 \% methyl cellulose solution used in RT-DC measurements, in contrast, features a much lower value of $R_\eta = \num{18.7\pm0.4}$.

\begin{table}[H]
	\centering
		\begin{tabular}{l|p{3cm}|p{1.5cm}|p{1.5cm}|p{1.5cm}|p{2cm}}
%			Author & Material & $\eta_0/\unit{\pascal\second}$ & $\eta_\infty/\unit{\pascal\second}$ & $R_\eta$ & extracted from \\
			Author & Material & $\eta_0/Pas$ & $\eta_\infty/Pas$ & $R_\eta$ & extracted from \\
			\hline
			Amorim\cite{amorim2021} & pre-crosslinked alginate &  >1e4 & <3e-1 & >3e4 & Fig. 3b \\
			\rowcolor{secondaryBackground}
			Pössl\cite{possl2021} & blend & 3e2 & 1e-4 & 3e6 & Table 4 \\
%			Paxton\cite{paxton2017} &Nivea Creme\n Alginate& 1e5\n5e4 & <1\n<5e1 & >1e5\n>1e3 & Fig. 4b \\
			Paxton\cite{paxton2017} & Alginate& 5e4 & <5e1 & >1e3 & Fig. 4b \\
			\rowcolor{secondaryBackground}
			O'Connell\cite{oconnell2020} & pluronic F-127 & >1e6 & <8 & > 2e5 & Chapter 7 Fig. 13 \\
		\end{tabular}
	\caption{Compilation of literature data on typical bioinks showing that a very viscosity ratio $ R_ \eta $ is a generic feature of these liquids. 
	\SG{ cross-check these values }
	}
	\label{tab:bioinks}
\end{table}

@@

\SG{ make sure that all papers cited in the introduction as ''3D'' are discussed here as well }
\SG{ and vice versa: make sure that all papers discussed here are also cited in the introduction }
Since the early days of LBM variants for viscoelastic fluids have been developed. 
Keeping in mind our target geometries, we focus here on three dimensional formulations.
The first approaches avoided an explicit modeling of a viscoelastic constitutive equation and simply modified the LB collision operator to obtain a fluid obeying Jeffery's viscoelastic model \cite{lallemand2003}. \SG{ can we say more about Jeffery? Why is it bad? Is it shear-thinning? }
More advcanced approaches include the work of Onishi \textit{et al.} \cite{onishi2005, onishi2006} who considered a non-shear-thinning Oldroyd-B model using a microscopic Fokker-Planck equation for the polymer dumbbells. 
The work of \cite{wang2020} solved the advection part of the model using a second LBM running in parallel to the Navier-Stokes LBM.
The same technique was used by \cite{malaspinas2010} and \cite{ma2020} for the Oldroyd-B and FENE-P models.
Later, hybrid schemes coupling an LB solver for the solvent to a finite-difference \cite{gupta2015_2, gupta2015, gupta2016} or finite-volume \cite{kuron2021} scheme for the polymer stress (or, alternatively, the polymer conformation) tensor were introduced.
To improve numerical stability, some authors propose the inclusion of an artificial diffusivity \cite{malaspinas2010,ma2020,wang2020,wang2021}, which introduces a small error compared to the analytical solution of the constitutive equation. 
Other remedies include special decompositions of the stress tensor\cite{dzanic2022} or optimization of the solvent contribution\cite{gupta2015} \SG{ rephrase and give some more details }.  
Coupling of the resulting polymer stress to the Navier-Stokes LBM can be achieved via the inclusion of point forces or via a modification of the equilibrium populations.
A recent comparison in 2D found that the former in general leads to better stability \cite{dzanic2022}.

The stable range of Weissenberg numbers covered by existing methods is up to $ \mathrm{Wi}  \approx  10$ for the method of \cite{malaspinas2010} and up to $ \mathrm{Wi}  \approx 100$ in \cite{gupta2015,ma2020,gupta2016}.
However, another stability criterion, which is often overlooked is the viscosity ratio $ R_ \eta $ as defined in eq.~(\ref{eq:R_eta}). 
Existing methods often stay at or below $R_\eta = 1$ \cite{gupta2015,onishi2006,wang2020,su2013_2,gupta2015_2}, while others report being limited below $R_\eta = 10$ \cite{malaspinas2010,ma2020,kuron2021} due to stability issues. 
The details are listed in table~\ref{tab:viscRatio}.
Comparing to the rheological data presented in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc}, we conclude that existing LB methods appear to be appropriate only for relatively dilute solutions and in particular do not cover the viscosity ratio required by technologically important cell carrier fluids such alginate and methyl cellulose.
We therefore set out to develop a novel viscoelastic LBM scheme which we will present in the next section.

@@

The Lattice-Boltzmann method (LBM) has become a standard tool in numerical fluid mechanics \cite{Kruger_book} and we therefore describe it here only very briefly.
We use a D3Q19 lattice in full three dimensions together with a single relaxation time collision operator.
Our implementation is based on FluidX3D which has been extensively used and validated in earlier publications \cite{Lehmann.20229da}.
The core of the algorithm is implemented in OpenCL and thus capable of running on GPUs from different vendors.
Boundaries are modeled using the half-way bounce back algorithm. 
\SG{ and inflow/outflow: give details }

Physical units are mapped to lattice units to obtain a dimensionless relaxation time of 1.
The employed lattice dimensions vary for the different applications and are specified below. 
To drive the flow, we use either a constant body force (modeling a physical pressure gradient) or inflow-outflow boundaries as specified in the different application scenarios below.

@@

\section{Rheology measurements of alginate solutions}
\label{app:experiment}
